\documentclass[conference]{IEEEtran}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{oreboot}

\author{\IEEEauthorblockN{Ronald G Minnich}
\IEEEauthorblockA{\textit{Google} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{Ryan O'Leary}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}

}

\maketitle

\begin{abstract}
Firmware is the software that first runs on CPUs when they are turned on. 
The term "firmware" arises from the fact that it is stored in memory which maintains
its contents across power cycles, so-called non-volatile memory (NVRAM). 
Because of its crucial place in startup, its proprietary nature, and the difficulty in changing it, firmware has proven to be an effective place to store security exploits\cite{cui2013firmware}. Many of these exploits are made easier
by simple programming practice errors not caught by C compilers\cite{wojtczuk2015attacks}.

In this paper we present oreboot, a new open source, portable firmware system. 
oreboot is written in Rust\cite{balasubramanian2017system}, a new language which 
has properties designed to avoid the problems associate with assembly language and C.

oreboot is now working on RISCV, ARM, and x86 CPUs, in both 32- and 64- bit implementations. In this paper, we discuss oreboot's architecture, target implementations,
and lessons learned from writing firmware Rust. 


\end{abstract}

\begin{IEEEkeywords}
oreboot, firmware, coreboot
\end{IEEEkeywords}

\section{*Introduction}
\subsection{*The nature of firmware}
The software that runs on hardware is always loaded by software already running on the hardware. 
It is reasonable to ask: how does this work at the start? What software chicken lays the first software egg, from which grow more software-laying chickens?
This fundamental question is rarely asked, rarely answered, and little understood; yet it lies at the heart of how our computers work.
It has also changed very little since the earliest computers.
For purposes of discussion, we are going to exclude systems created before 1965, i.e. we will discuss only "modern" systems.
Computers at power-on/reset (usually called POR) need to have something to run, and that is where firmware comes in. Firmware is made accessible to the CPU so it can load another, more complex program from a disk, magnetic tape, paper tape, or other media. Rarely, this more complex program is a full end-user operating system; most commonly it is another step in the boot sequence\footnote{there are exceptions, e.g. \cite{minnich2000linux}.}

This process, in similar form, plays out on every computer system we use today. But the question remains: what starts the process off? 
In general, there are two main techniques to provide firmware at Power-On/Reset:
\begin{enumerate}
    \item a small auxiliary processor loads firmware into memory or L1 cache, then prepares the main CPU to run it. This technique is used in mainframes, some minicomputers, the DEC Alpha, and modern AMD chips. 
\item a nonvolatile memory, sometimes readonly, contains a program that loads a more complex program; the CPU immediately starts running this code at Power-On/Reset.
\end{enumerate}
The first technique requires a processor to load the firmware; that processor in turn has firmware. In one system, IBM's Power 8, Power 8 firmware is loaded by a Power workstation. The firmware for the CPU in that workstation is loaded by a smaller Power PC CPU. That CPU loads its firmware from a memory-addressable FLASH part. 

In all cases, when we chase the source of firmware down to its origin, there is always a memory-addressable part somewhere in this chain containing firmware for a CPU.
It is not turtles all the way down; at some point there is a bottom turtle, resting on a solid foundation of non-volatile memory such as FLASH. 
This paper concerns what firmware  \textit{all} those CPUs run, be it loaded via a smaller CPU, or contained in a non-volatle memory. 

\subsection{*A typical firmware boot sequence: modern ARM CPUs}
We close the discussion of firmware booting with a look at a typical boot sequence from an ARM 32-bit CPU, starting at Power-On/Reset

\begin{itemize}
    \item At Power-On/Reset the CPU loads the Instruction Pointer with physical address 0, and executes from Read-Only Memory (ROM) code (Boot Level 0, or BL0) stored in the chip itself. BL0 optionally verifies and loads code (BL1) from FLASH into on-chip SRAM.
    \item The BL1 copies a BL2 from FLASH to on-chip SRAM.
    \item The larger, more complex BL2 gets memory working, and copies a BL3 into memory.
    \item The BL3, which nowadays has up to four components, loads more parts of itself into memory.
    \item The BL3, having found a boot medium, loads a \textit{boot block} from it.
    \item The boot block is a direct analog of the BL0, in that it is small and exists only to load a more complex program, this time from the boot medium.
    \item The next, more complex program locates a kernel and loads it.
    \item The kernel runs and, in some cases, fetches a kernel, and replaces itself with it.
\end{itemize}

This process is actually greatly simplified from 64-bit ARM and x86 CPUs. Those systems contain more verification and hence more code. But the basic sequence is similar.

\subsection{*oreboot}

how it is implemented 
security issues 
open source firmware projects
coreboot
oreboot as a downstream fork of coreboot\cite{minnich2000linux}
Open-source and GPLv2
Rust
Absolutely no C code.
Small pieces of assembly where necessary (ex: initializing stack pointer)
Coreboot assembly code is very useful for these tricky bits
Jump to kernel as quickly as possible
Firmware contains no network stack, disk drivers, debug shells, …
Those features are provided by payloads such as LinuxBoot
Strict policy for accepting closed-source blobs
Only an issue for the x86 port
Current RISC-V ports are fully open-source


\section{*Survey of Current Open-Source Firmware}

U-boot (1999-)
LinuxBIOS (1999-2008)
Coreboot (2008-)
NERF (2016-)
Linuxboot (2017-)
u-bmc (2018-)
SlimBoot (2018-) [sort of, it’s a UEFI DXE]

-- Too much C and assembly code.

\section{*Language Choice}

\subsection{*Rust for Firmware}

In recent decades, we have seen a flourish of programming languages and great improvements to language design which provide better safety features and ease of development. Many of these new languages such as Go, Python, Java, ... are unsuitable for the lowest layer of firmware.

\begin{itemize}
    \item Dependence on OS environment (such as syscall interface, memory managment)
    \item Runtime (GC, green threads)
\end{itemize}

Yes there are non-standard "reduced" implementations (such as tinygo, micropython, ...) of these languages, but they are non-standard / non-mainstream. For this reason, C still dominates firmware. All mainstream C compilers (such as GCC, clang) will allow you to build software in a standalone mode suitable for firmware.

Rust was originally designed by MOzilla in 200x for writing web browsers. The goal of making them more performant and secure. Today, it is used in kernals (see Tock \cite{b1} and Redox \cite{b3}), embedded firmware (see ??), usermode applications (see ??), UEFI drivers (see ??) and elsewhere.

\subsection{*Rust Safety}

The Rust compiler offers a number of compile-time safety features not present in other languages. In this section, three features we found useful for firmware are highlighted:

\begin{itemize}
    \item Borrow Checker
        \begin{verbatim}
for i in xs {
  print(i);
  xs.push(7); // compile error since xs is immutably borrowed
}
        \end{verbatim}
    \item Memory-safe concurrency
    \item Printf \begin{quote}format!("Rust {} {}", 4, "firmware");\end{quote}
\end{itemize}

\section{*Architecture}

\subsection{*Boot Flow}

\begin{itemize}
    \item \textbf{Boot Blob}: Executes directly from flash
First instruction
Initialize CPU
Debug UART print “Welcome to Oreboot”
Setup SRAM/CAR
Find and jump to Rom Stage
A fair bit of assembly code
    \item \textbf{Rom Stage}: Executes directly from flash
Has very little ~30KiB-8MiB of SRAM/CAR. Not enough for Linux yet!
Initialize RAM
    \item \textbf{Payloader stage}: Has one job
Find, load and run a payload
The payloader has no drivers, storage drivers, USB stack, etc… This is a big complexity reduction compared to your classic coreboot.
    \item \textbf{LinuxBoot}: Linux + Initramfs (or another kernel of your choosing)
Kernel (not oreboot) can optionally load another kernel from the disk or network and kexec
    \item \textbf{OS}
\end{itemize}

\subsection{*Driver Model}

\begin{verbatim}
pub type Result<T> = core::result::Result<T, &'static str>;
pub const EOF: Result<usize> = Err("EOF");
pub const NOT_IMPLEMENTED: Result<usize> = Err("not implemented");

pub trait Driver {
	/// Initialize the device.
	fn init(&mut self) -> Result<()>;
	/// Positional read. Returns number of bytes read.
	fn pread(&self, data: &mut [u8], pos: usize) -> Result<usize>;
	/// Positional write. Returns number of bytes written.
	fn pwrite(&mut self, data: &[u8], pos: usize) -> Result<usize>;
	/// Shutdown the device.
	fn shutdown(&mut self);
}
\end{verbatim}

\section{*Use Cases}

\subsection{*RISCV - OpenTitan}

RISC-V rv32imc
Embedded, not Linux capable
Open-source design for a root-of-trust
No ASIC available yet
Oreboot current runs on:
Verilator
FPGA (Artix-7)
Currently trying to boot Tock kernel

\subsection{*ARM - u-BMC}

ARM11
BMC Platform
Open-source memory initialization code
Converted to .rs with a Go program
Bootblob
1. Initialize CPU
2. Initialize Bus
3. Initial UART Print
4. Initialize SRAM
Romstage
5. Memory init


\subsection{*x86}

We prefer all pieces of the firmware to be open-source; but can accept an ME and FSP binary blob for x86 architectures.
Blobs must be essential to boot the system and not provide any extraneous functionality which could not be implemented in Oreboot.
Blobs must be redistributable and are ideally available on GitHub.
Blobs must not be submitted to github.com/oreboot/oreboot. We prefer blobs to be submitted to github.com/oreboot/blobs, github.com/coreboot/blobs or some other GitHub repository.
The blobs must be in a binary format. No additional C code, assembly files or header files are acceptable.
Any compromises to the language safety features of Rust must be explicitly stated.


\section*{Acknowledgment}

TODO

\section*{References}
\bibliographystyle{unsrt}
\bibliography{oreboot}


\vspace{12pt}

\end{document}
